Associahedra for finite type cluster algebras \\ and minimal relations between $\b{g}$-vectors

### 2019-07-09 thirdPaper/draft/typeConeAssociahedra.tex
edditing structure of the paper

\section*{Main results and logical dependencies}

We give a brief overview of our main results.
Their logical dependencies are recalled in a schematic diagram at the end of this section.

Let~$\Fan$ be an essential, complete, simplicial, polyhedral fan in~$\R^n$ (all definitions are recalled in \cref{ssec:polytopes and fans}).
The type cone~$\typeCone(\Fan)$ of~$\Fan$ (\cref{ssec:type cone}), introduced by P. McMullen, is a device that encodes the information about all possible polytopal realizations of~$\Fan$.
If the simplicial fan~$\Fan$ has the unique exchange relation property (\cref{def:uerp}), then its type cone is easier to describe and study.
In particular, if moreover the type cone~$\typeCone(\Fan)$ is simplicial (\cref{rem:dimTypeCone}), then all possible polytopal realizations of~$\Fan$ can be described explicitely in~$\R^N_{>0}$, where~$N$ is the number of rays of~$\typeCone(\Fan)$, in terms of~$N-n$ positive real parameters (\cref{coro:simplicialTypeCone}).
This gives a new point of view on~\cite{ArkaniHamedBaiHeYan,BazierMatteDouvilleMousavandThomasYildirim}, illustrated in \cref{subsec:typeConeAsso} for classical associahedra, that we then apply to two families of~$\b{g}$-vector fans generalizing the~$\b{g}$-vector fans of classical associahedra: The~$\b{g}$-vector fans of cluster algebras of finite type (generalized associahedra), and the~$\b{g}$-vector fans of gentle algebras (non-kissing associahedra, non-crossing associahedra, generalized accordiohedra).
We also describe the type cone for graph associahedra (\cref{subsec:typeConeGA}), to which our strategy does not apply.

@@

We first consider the~$\b{g}$-vector fan~$\Fan(\B_\circ)$ of a skew-symmetrizable cluster algebra of finite type, with respect to any initial exchange matrix of finite type~$\B_\circ$.
A reformulation of~\cite[Thm 7.5]{BuanMarshReinekeReitenTodorov} gives the first step towards applying our type cone strategy.

@@

The second step follows from our main result in \cref{sec:clusterCategories}, which makes use of representation theory of finite dimensional algebras and additive categorification of cluster algebras.

@@

We consider the set~$\meshes$ of all pairs~$\{x,x'\}$ related by non-initial mesh mutations (\cref{def:meshMutation}), and we denote by~$\principalVariables$ the set of cluster variables.
Let~$\{x,x'\}\in\meshes$, and let~$X,X'$ be two clusters such that~$X\setminus\{x\}=X'\setminus\{x'\}$.
For any~$y\in\principalVariables$, define~$\alpha_{x,x',y}$ to be~$|b_{xy}|$ if~$y\in X\cap X'$ and 0 otherwise.
By the unique exchange relation property, the~$\alpha_{x,x',y}$'s do not depend on the specific choice of the clusters~$X,X'$.

@@

\noindent \cref{prop:exchangeablePairsNKC} (ii): \emph{The non-kissing fan~$\Fan(\quiver)$ has the unique exchange relation property.}

@@

Write~$\strings$ for the set of all strings of~$\quiver$, and~$\walks$ for the set of all walks on~$\quiver$ (i.e. of all strings of the blossoming bound quiver of~$\quiver$ that join two blossom vertices).
If~$\sigma$ is a string of~$\quiver$, we make use of the evocative notations~$\cc{\sigma}, \hh{\sigma}, \hc{\sigma}, \ch{\sigma}$ for the walks obtained by adding, in the blossoming bound quiver, hooks or cohooks at each end of~$\sigma$.
We make use of our type cone approach to deduce:

@@

@@

In \cref{sec:clusterCategories}, we prove an analogue for cluster categories of M. Auslander's result~\cite{Auslander1984}.
We are mainly motivated by the fact that itt has for a consequence that the type cones of the cluster fans of finite type are simplicial.
However, this result is of independent interest and sheds new lights on the Grothendieck groups of cluster categories.

@@

Schematic diagram:\yann{Vincent, Arnau, je suis certain que vous feriez bien mieux que moi. Puis-je vous demander de le représenter ?}

I.1 explains the type cone approach to the problem

This strategy is used in I.2.2 for cluster algebras and in I.2.3 for gentle algebras.

II.3 cluster categories (2-Calabi--Yau triangulated categories) is used in 1.2.2 for proving U.E.R.P. and simpliciality of the type cone.

II.4 generalises II.3, is used in I.2.3 for proving the U.E.R.P. and offers a different proof of simpliciality of the type cone .

@@

\begin{corollary}
Consider a graph~$\graphG$ on~$\ground$ with tubes~$\tubes$ and a height vector~$\b{h} \in \R^{\tubes}$ such that~$\b{h}_{\varnothing} = \b{h}_{\graphG} = 0$ and satisfying 
 $\b{h}_{\tube[s] \ssm \{v'\}} + \b{h}_{\tube[s] \ssm \{v\}} > \b{h}_{\tube[s]} + \b{h}_{\tube[s] \ssm \{v, v'\}}$ for any tube~$\tube[s] \in \tubes$ and distinct non-disconnecting vertices~$v,v'$ of~$\tube[s]$. 
Then the nested fan~$\gvectorFan[\graphG]$ is the normal fan of the graph associahedron
\[
\set{\b{x} \in \R^\ground}{\dotprod{\gvector{\tube}}{\b{x}} \le \b{h}_{\tube} \text{ for any tube } \tube \in \tubes}.
\]
\end{corollary}

@@

\begin{proposition}
The type cone~$\typeCone \big( \gvectorFan[\graphG] \big)$ is simplicial if and only if~$\graphG$ is a path.
\end{proposition}

>>>>

% \section*{Main results and logical dependencies}
\subsection*{Structure of the paper: main results and logical dependencies}

We give a brief overview of our main results and the structure of the article. In particular, some of the geometric realizations in~\cref{part:geometry} depend on algebraic results whose proof is delayed until~\cref{part:algebra}. The goal of this section is to clarify their logical dependencies. 
% Their logical dependencies are recalled in a schematic diagram at the end of this section.

Let~$\Fan$ be an essential complete polyhedral fan in~$\R^n$ (all definitions are recalled in~\cref{ssec:polytopes and fans}).
The type cone~$\typeCone(\Fan)$ of~$\Fan$ (\cref{ssec:type cone}), introduced by P. McMullen~\cite{McMullen-typeCone}, is a polyhedral cone that parametrizes the set of all possible polytopal realizations of~$\Fan$. These realizations can be described explicitly as affine sections of the nonnegative orthant~$\R^N_{\geq0}$, where~$N$ is the number of rays of~$\Fan$ (\cref{prop:alternativePolytopalRealization}).  
If $\Fan$ is a simplicial fan and has the unique exchange relation property (\cref{def:uerp}), then its type cone is easier to describe and study.
And if moreover the type cone~$\typeCone(\Fan)$ is simplicial (\cref{rem:dimTypeCone}), then all possible polytopal realizations of~$\Fan$ can be described explicitely in terms of~$N-n$ positive real parameters (\cref{coro:simplicialTypeCone}).
This gives a new point of view on~\cite{ArkaniHamedBaiHeYan,BazierMatteDouvilleMousavandThomasYildirim}, illustrated in \cref{subsec:typeConeAsso} for classical associahedra, that we then apply to two families of~$\b{g}$-vector fans generalizing the~$\b{g}$-vector fans of classical associahedra: The~$\b{g}$-vector fans of cluster algebras of finite type (generalized associahedra) in~\cref{subsec:typeConeCA}, and the~$\b{g}$-vector fans of gentle algebras (non-kissing associahedra, non-crossing associahedra, generalized accordiohedra) in~\cref{subsec:typeConeNKC}.
We also completely describe the facets of the type cone for graph associahedra in~\cref{subsec:typeConeGA}, to which our strategy does not apply, as only the classical associahedron has a simplicial type cone.

@@

We first consider the~$\b{g}$-vector fan~$\Fan(\B_\circ)$ of a skew-symmetrizable cluster algebra of finite type, with respect to any initial exchange matrix of finite type~$\B_\circ$. We provide the following polytopal realizations of~$\Fan(\B_\circ)$:
% We consider the set~$\meshes$ of all pairs~$\{x,x'\}$ related by non-initial mesh mutations (\cref{def:meshMutation}), and we denote by~$\principalVariables$ the set of cluster variables.
% Let~$\{x,x'\}\in\meshes$, and let~$X,X'$ be two clusters such that~$X\setminus\{x\}=X'\setminus\{x'\}$.
% For any~$y\in\principalVariables$, define~$\alpha_{x,x',y}$ to be~$|b_{xy}|$ if~$y\in X\cap X'$ and 0 otherwise.
% By the unique exchange relation property, the~$\alpha_{x,x',y}$'s do not depend on the specific choice of the clusters~$X,X'$.

Here, $\meshes$ represents the set of all pairs~$\{x,x'\}$ related by non-initial mesh mutations (\cref{def:meshMutation}), and~$\principalVariables$ the set of cluster variables.
For $\{x,x'\} \in \meshes$ and $y \in \principalVariables$, the coefficient $\alpha_{x,x',y}$ is defined as follows: $\alpha_{x,x',y}$ is set to be~$|b_{xy}|$ if~$y\in X\cap X'$ and $0$ otherwise, where~$X,X'$ are any two clusters such that~$X\setminus\{x\}=X'\setminus\{x'\}$. By the unique exchange relation property, the~$\alpha_{x,x',y}$'s do not depend on the specific choice of the clusters~$X,X'$.

Some crucial pieces of the proof of~\cref{thm:allPolytopalRealizationsCA} use cluster categories (2-Calabi--Yau triangulated categories) and are deferred to~\cref{sec:clusterCategories}. Namely, a reformulation of~\cite[Thm 7.5]{BuanMarshReinekeReitenTodorov}, proved in~\cref{sec:clusterCategories}, gives the first step towards applying our type cone strategy.

@@

And the second step follows from our main result in \cref{sec:clusterCategories}, which makes use of representation theory of finite dimensional algebras and additive categorification of cluster algebras.

@@

\medskip
\bigskip

@@

\noindent \cref{prop:exchangeablePairsNKC} (ii): \emph{The non-kissing fan~$\Fan(\quiver)$ has the unique exchange relation property.}
\arnau{In the proof of \cref{prop:exchangeablePairsNKC} it says that the UERP follows from PPP17}
\medskip

@@

Combining these pieces, we make use of our type cone approach to deduce:

@@

Here, $\strings$ is the set of all strings of~$\quiver$, and~$\walks$ the set of all walks on~$\quiver$ (i.e. of all strings of the blossoming bound quiver of~$\quiver$ that join two blossom vertices). If~$\sigma$ is a string of~$\quiver$, we make use of the evocative notations~$\cc{\sigma}, \hh{\sigma}, \hc{\sigma}, \ch{\sigma}$ for the walks obtained by adding, in the blossoming bound quiver, hooks or cohooks at each end of~$\sigma$.

\noindent {\bf Graphical nested complexes} (\cref{subsec:typeConeGA})

Even though graph associahedra do not have simplicial type cones except for the path (\cref{prop:simplicialnestedfan}), we provide a non-redundant characterization of the type cones of graphical nested fans:

\noindent \cref{cor:typeconenestedfan}:\emph{
Consider a graph~$\graphG$ on~$\ground$ with tubes~$\tubes$. 
Then, for a height vector~$\b{h} \in \R^{\tubes}$, the nested fan~$\gvectorFan[\graphG]$ is the normal fan of the graph associahedron
\[
\set{\b{x} \in \R^\ground}{\dotprod{\gvector{\tube}}{\b{x}} \le \b{h}_{\tube} \text{ for any tube } \tube \in \tubes}
\]
if and only if~$\b{h}_{\varnothing} = \b{h}_{\graphG} = 0$ and
 $\b{h}_{\tube[s] \ssm \{v'\}} + \b{h}_{\tube[s] \ssm \{v\}} > \b{h}_{\tube[s]} + \b{h}_{\tube[s] \ssm \{v, v'\}}$ for any tube~$\tube[s] \in \tubes$ and distinct non-disconnecting vertices~$v,v'$ of~$\tube[s]$.
}

@@

In \cref{sec:clusterCategories}, we prove an analogue for cluster categories of M. Auslander's result~\cite{Auslander1984}. \arnau{Which result? Is it famous? Should we say something like Auslaunder's famous result from...? }
We are mainly motivated by the fact that it has for a consequence that the type cones of the cluster fans of finite type are simplicial.
However, this result is of independent interest and sheds new lights on the Grothendieck groups of cluster categories.

@@

% Schematic diagram:\yann{Vincent, Arnau, je suis certain que vous feriez bien mieux que moi. Puis-je vous demander de le représenter ?}
% 
% I.1 explains the type cone approach to the problem
% 
% This strategy is used in I.2.2 for cluster algebras and in I.2.3 for gentle algebras.
% 
% II.3 cluster categories (2-Calabi--Yau triangulated categories) is used in 1.2.2 for proving U.E.R.P. and simpliciality of the type cone.
% 
% II.4 generalises II.3, is used in I.2.3 for proving the U.E.R.P. and offers a different proof of simpliciality of the type cone .

@@

\begin{corollary}\label{cor:typeconenestedfan}
Consider a graph~$\graphG$ on~$\ground$ with tubes~$\tubes$. 
Then, for a height vector~$\b{h} \in \R^{\tubes}$, the nested fan~$\gvectorFan[\graphG]$ is the normal fan of the graph associahedron
\[
\set{\b{x} \in \R^\ground}{\dotprod{\gvector{\tube}}{\b{x}} \le \b{h}_{\tube} \text{ for any tube } \tube \in \tubes}
\]
if and only if~$\b{h}_{\varnothing} = \b{h}_{\graphG} = 0$ and
 $\b{h}_{\tube[s] \ssm \{v'\}} + \b{h}_{\tube[s] \ssm \{v\}} > \b{h}_{\tube[s]} + \b{h}_{\tube[s] \ssm \{v, v'\}}$ for any tube~$\tube[s] \in \tubes$ and distinct non-disconnecting vertices~$v,v'$ of~$\tube[s]$.
\end{corollary}

@@

\begin{proposition}\label{prop:simplicialnestedfan}
The type cone~$\typeCone \big( \gvectorFan[\graphG] \big)$ is simplicial if and only if~$\graphG$ is a path.
\end{proposition}
